---
title: 'Module 1: Introduction to R'
author: "Jeroen Claes"
output:
  revealjs::revealjs_presentation:
    css: white.css
    self_contained: no
    transition: none
  html_document:
    toc: no
  word_document:
    toc: no
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.height = 4, dpi = 300)
library(pander)
library(knitr)
library(tibble)
```


## Introduction

## About this course (1/3)
- Four goals:
    - Introduction to R (This module)
    - Viewing, manipulating, and exporting tabular data (Module 2)
    - Exploring qualitative variables (Module 3)
    - Exploring the relationships between qualitative variables (Module 4)
    - Elements of logistic regression (Module 5)

## About this course (2/2)
- Method:
    - Code demonstrations/discussion. We will use datasets from (my) earlier research for the demonstrations 
    - You will perform the exercises in a controlled environment on http://www.jeroenclaes.be/statistics_for_linguistics. The links to the labs will be on the slides
    - You will perform the Module-final projects with your *own dataset* on your own machine. 
    - The Module-final projects add up to a completely finished script that answers the research question you brought with you
- You'll walk away with: 
    - A solid understanding of how to load, manipulate, explore and report on data with R
    - A finished script that analyzes the research question you defined 


## In this module
- A general introduction
- A general introduction to R
- Programming 101
- Importing tabular data into R


## A general introduction to R

## What is R? 
- R's full name is: *R: A language and environment for statistical computing and graphics*
- Open-source programming language designed specifically for the manipulation, analysis, visualisation, and modeling of data  
- R was primarily designed for interactive use, for which it includes many features that make your life easier (but not necesarily your computer's)

## What is RStudio? (1/2)
- RStudio is THE company in the R ecosystem 
- RStudio develops software to support the use of R in enterprise, among others:
    - RStudio IDE (Integrated Development Environment, a piece of software that helps you write R code and analyse data with R)

## What is RStudio (2/2)
- RStudio IDE makes programming in R easier:
    - Hints packages
    - Hints function names
    - Hints objects in memory
    - Hints column names
    - Allows you to have your code and the window that executes your code side-by-side
- The course package contains a handy cheat sheet that explains all the options and possibilities of the RStudio IDE (courtesy of RStudio, Inc.)
    
## Let's get our hands dirty with code!
- From this point on, you should try to execute the code snippets I type on the screen

## Packages (1/3)
- RStudio also develops a lot of **packages**, extensions for the R language that extend its functionality  
- At the end of January 2019, there were **13,626 packages** available on *The Comprehensive R Archive Network (CRAN)* (± 1,000 more than a year earlier)

## Packages (2/3)
- To install a *package*,  type in `install.packages("NameOfThePackage")`. You only have to do this once
- To load a package, you have to type `library(NameOfThePackage)` once per session
- To consult the documentation of a particular function, simply type in `?NameOfTheFunction`, AFTER loading the package

## Packages (3/3)
- The following code will install the lightweight `RColorBrewer` package, which is used to create color palettes for plots

```{r documentation, echo=TRUE, eval=FALSE}
install.packages("RColorBrewer")
library(RColorBrewer)
?brewer.pal
RColorBrewer::brewer.pal()
```

## Taking stock
- In this part we have learned:
    - How to install packages
    - How to load packages
    - How to consult the documentation of functions

## Programming 101

## Programming 101: Objects and variables (1/4)
- Objects: a piece of information in R memory
- Variables: a reference to this information
- In the example 
  - *hello* is an object
  - *a* is a variable

```{r cars, echo = TRUE}
a <- "hello"
```

## Programming 101: Objects and variables (2/4)
- You can create variables (and objects) by typing in a name and assigning a value to that name with the assignment operator `<-`
    
## Programming 101: Objects and variables (3/4)
- There's a few things you have to keep in mind when assigning variables:
    - Variables are case sensitive: *ab* vs. *Ab* vs. *aB*
    - Spaces and non-ASCII characters (e.g., *hé* ) are not allowed in variable names
    - Variable names cannot start with a number (e.g., `3b`)
    - **BUT** if you wrap your variable names with backticks *``*, all of this **IS** allowed
    
## Programming 101: Objects and variables (4/4)
- You can save yourself a lot of frustration if you adopt a naming system and stick to it
- There's two commonly used systems:
    - *snake_case*
    - *camelCase*

## Programming 101: using R as a calculator 
- Because variables refer to objects, you can easily perform calculations with numeric variables

```{r rekenmachine, echo = TRUE}
a <- 2
b <- 4
a*b
```

## Programming 101: Vectors: the basic data structure in R (1/8)
- Up until now, we have only considered variables that store just one piece of information (e.g., a single number, *1*)
- Most often, we will deal with variables that contain multiple bits of information, e.g., a list of values such as *a, b, c*
- In R, these are called `vectors`. In fact, everything in R is a vector
- The types of variables we have considered up until now (scalars) are just vectors of length 1!

## Programming 101: Vectors: the basic data structure in R (2/4)
- The basic vector-creating function in R is `c()` (from **c***ombine*)
- R automatically guesses the data type of the vector

```{r vectors}
fruits <- c("apples", "oranges", "bananas")

numbers <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
```

## Programming 101: Vectors: the basic data structure in R (3/4)
- Since the basic data structure in R is the vector, most functions/operations (e.g., multiply by 2) apply equally well to a single element as they do to a vector of elements

```{r vectorizedOps}
numbers <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
numbers *2
numbers -2
numbers +2
```

## Programming 101: Vectors: the basic data structure in R (4/4)
- It is generally **unnecessary** to write loops in R to iterate over the different elements of a vector to get things done
- There's a first exercise waiting for you at: http://www.jeroenclaes.be/statistics_for_linguistics/class1.html

## Reading data into R | Importing CSV and Excel files 

## CSV files: a necessary evil (1/3)
- CSV (comma-separated values) files are a widely used data format to exchange datasets between platforms and systems
- CSV files are very useful, but they can also be annoying:
    - Fields that are not separated properly
    - Delimiter characters that are used within fields without quoting
    - Decimal separators that vary from locale to locale
    - Encoding issues (the translation between bits and human-readable characters)

## CSV files: a necessary evil (2/3)
- The last one is especially common, as there's hundreds of encoding options out there, depending on the user's locale
- Just some quick examples of the variety of encodings that are out there:
    - On Windows, Excel will produce CP-1252 or ISO-8559-1 encoded CSV files
    - Older versions of Excel for Mac would produce Mac-Roman encoding
    - Newer versions of Excel for Mac use UTF-8 (Unicode)
    
## CSV files: a necessary evil (3/3)
- Out of the box, R assumes the cross-platform UTF-8 encoding standard
- This means that if you read a ISO-8559-1-encoded CSV file without specifying the encoding, all non-ASCII characters (e.g., `é`) will look weird
- Let us first take a look at how you can load a UTF-8 encoded CSV file, keeping the encoding issue for later

## The `readr` package (1/2)
- R has build-in CSV readers (the functions `read.csv`, `read.csv2`, `read.tsv` and `read.delim`), but these are slow, inflexible and not fault-tolerant
- It is better to use the `readr` package, which is much faster, more flexible, and more fault-tolerant
```{r installReader, eval=FALSE}
install.packages("readr")
```

## The `readr` package (2/2)
- The `readr` package has four main reader functions, which accept the same set of arguments:
    - `read_csv`: read comma-separated files
    - `read_csv2`: read semicolon-separated files
    - `read_tsv`: read tab-separated files
    - `read_delim`: read files separated by an arbitrary delimiter character, which you can specify yourself

## How to read CSV-files with `readr` (1/3)
- Let us explore the syntax of the `readr` functions by looking at how we can read a dataset from the web

```{r readCodeFile}
library(readr)
dataSet <- read_csv("http://www.jeroenclaes.be/statistics_for_linguistics/datasets/class3_claes_2017.csv")
```

## How to read CSV-files with `readr` (2/3)

```{r readCodeFileEx, eval=TRUE, echo=F}
library(dplyr)
glimpse(dataSet)
```

## Dealing with encoding (and other locale issues) (1/4)
- Sometimes you will receive a file from a colleague and it's impossible to know what encoding it is in
- For cases like that, `readr` includes a function (`guess_encoding`) that determines the most likely encoding of a file
- Here we find that the most likely encoding is *ASCII* (which is a subset of UTF-8, the default), so we don't have to specify it explicitly

```{r guessencoding}
guess_encoding("http://www.jeroenclaes.be/statistics_for_linguistics/datasets/class3_claes_2017.csv")
```
    
## Dealing with encoding (and other locale issues) (2/4)
- When you do want to specify the encoding of a file, you can do so by defining a **locale** for your file
- Think about a locale as a complete specification of all culture/country-specific aspects of a file:
    - date names
    - date formatting
    - time formatting
    - decimal mark
    - thousand separator
    - time zone
    - encoding
- Of course you do not have to specify all of these aspects to change just the encoding, but know that these options are available

## Dealing with encoding (and other locale issues) (3/4)
- To change the locale for your file, you can assign the result of the `locale` function to the `locale` argument of the `readr` functions
- The following ficticious example would read a ISO-8859-1 file, in which commas are used as decimal separators, and full stops are used as thousand separators
```{r fileWithBadEncoding, eval=FALSE}
dataSet <- read_csv2("someFile.csv", locale=locale(encoding="ISO-8859-1", decimal_mark = ",",grouping_mark = "."))
```

## Taking stock 
- In this part we have learned:
    - How to import different types of CSV files into R, using functions provided by the `readr` package
    - How to detect the encoding of a CSV file
    - How to specify the locale settings of a CSV file

## Reading data into R: Excel files
- While CSV files are common, most organizations have much data sitting in Excel files
- We can import the contents of Excel files to R as well
- To this end, we can use the `readxl` package
```{r installreadExcel, eval=FALSE}
install.packages("readxl")
```

## The `readxl` package (1/4)
- For most use cases, the `read_excel` function will do everything you need to do
- It automatically detects the difference between `.xls` and `.xlsx`, based on the file extension
- If you want to specify the file type yourself (e.g., when the extension is wrong), use the functions `read_xlsx` and `read_xls` accordingly

## The `readxl` package (2/4)
- The following line will import the first sheet from the `sales.xlsx` file in your `demonstration_data` folder

```{r readCodeXlsx, eval=TRUE}
library(readxl)
dataSet <- read_excel("datasets/corpus_futuro.xlsx", sheet = 1)
```


## Taking stock
- In this part we've learned:
    - How to read in Excel sheets from excel files (.xlsx, .xls) with the `readxl` package
    - How to specify the sheet we want to load to R memory
    - How to specify the data types of the columns in our sheet

## Module overview and module-final project

## Module overview 
- In this module, we have covered:
    - A general introduction to R
    - How to assign objects to variables
    - How to perform basic arithmetic with R
    - How to create and manipulate vectors
    - How to read CSV data into R
    - How to read Excel spreadsheets into R

## Module-final project
- Write a script that imports your dataset(s)