---
title: "Module 3: Exploring qualitative variables"
author: "Jeroen Claes | <jeroen.claes@kuleuven.be> | <jeroen@cropland.be>"
output: 
 html_document:
  self_contained: false 
---

```{r, include=FALSE}
library(testwhat)
tutorial::go_interactive()
```

## 1. Factors 
### 1.1 Factor levels

```{r ex="factors1", type="pre-exercise-code"}
Sys.setlocale(category = 'LC_ALL','en_US.UTF-8')
library(readr)
library(dplyr)
dataSet <- read_csv("http://www.jeroenclaes.be/r_crashcourse_for_linguists/datasets/class3_claes_2017.csv")
```

```{r ex="factors1", type="sample-code"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column state to a factor

# Print its levels

```

```{r ex="factors1", type="solution"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column state to a factor
dataSet<- dataSet %>% 
  mutate(state=as.factor(state))
# Print its levels
levels(dataSet$state)
```

```{r ex="factors1", type="sct"}
 test_output_contains('levels(dataSet$state)',  incorrect_msg  = "Make sure to print the levels of the factor!")
success_msg("Great!")
```

### 1.2 Re-organizing factors

```{r ex="factors3", type="pre-exercise-code"}
Sys.setlocale(category = 'LC_ALL','en_US.UTF-8')
library(readr)
library(dplyr)
dataSet <- read_csv("http://www.jeroenclaes.be/r_crashcourse_for_linguists/datasets/class3_claes_2017.csv")
```

```{r ex="factors3", type="sample-code"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column corpus to a factor

# Print its levels

# Relevel the corpus factor: its first value should be 'Twitter'

# Print its levels again

```

```{r ex="factors3", type="solution"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column corpus to a factor
dataSet <- dataSet %>%
  mutate(corpus=as.factor(corpus))
# Print its levels
levels(dataSet$corpus)
# Relevel the corpus factor: its first value should be 'Twitter'
dataSet <- dataSet %>%
  mutate(corpus=relevel(corpus, ref="Twitter"))
# Print its levels again
levels(dataSet$corpus)

```

```{r ex="factors3", type="sct"}
test_output_contains("levels(dataSet$corpus)", incorrect_msg = "Make sure to print the levels of the factor!")
success_msg("Great!")
```

## 2. Frequencies: counting values
### 2.1 Counting the `negation` variable 
```{r ex="counts_negation_freq", type="pre-exercise-code"}
Sys.setlocale(category = 'LC_ALL','en_US.UTF-8')
library(readr)
library(dplyr)
dataSet <- read_csv("http://www.jeroenclaes.be/r_crashcourse_for_linguists/datasets/class3_claes_2017.csv")
```

```{r ex="counts_negation_freq", type="sample-code"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column negation to a factor

# Compute a table for the negation variable, store it in the variable 'negationTab'

# Convert 'negationTab' to a data.frame

# Print 'negationTab'


```

```{r ex="counts_negation_freq", type="solution"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column negation to a factor
dataSet <- dataSet %>%
  mutate(negation=as.factor(negation))
# Compute a table for the negation variable, store it in the variable 'negationTab'
negationTab<-table(dataSet$negation)
# Convert 'negationTab' to a data.frame
negationTab<-as.data.frame(negationTab)
# Print 'negationTab'
negationTab
```

```{r ex="counts_negation_freq", type="sct"}
test_object("negationTab",incorrect_msg="Make sure to compute the table!")
test_output_contains("negationTab", incorrect_msg = "Make sure to print the table!")
success_msg("Great!")

```

### 2.2 Counting the `province` variable 
```{r ex="counts_province", type="pre-exercise-code"}
Sys.setlocale(category = 'LC_ALL','en_US.UTF-8')
library(readr)
library(dplyr)
dataSet <- read_csv("http://www.jeroenclaes.be/r_crashcourse_for_linguists/datasets/class3_claes_2017.csv")
```

```{r ex="counts_province", type="sample-code"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column province to a factor

# Compute a table for the province variable, store it in the variable 'provinceTab'

# Convert 'provinceTab' to a data.frame

# Print 'provinceTab'


```

```{r ex="counts_province", type="solution"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column province to a factor
dataSet <- dataSet %>%
  mutate(province<-as.factor(province))
# Compute a table for the province variable, store it in the variable 'provinceTab'
provinceTab<-table(dataSet$province)
# Convert 'provinceTab' to a data.frame
provinceTab<-as.data.frame(provinceTab)
# Print 'provinceTab'
provinceTab
```

```{r ex="counts_province", type="sct"}
test_object("provinceTab", incorrect_msg="Make sure to compute the table!")
test_output_contains("provinceTab", incorrect_msg="Make sure to print the table!")
success_msg("Great!")
```

## 3. Proportions
### 3.1 Proportions for the `Typical.Action.Chain.Pos` variable 
```{r ex="proportionsTypical.Action.Chain.Pos", type="pre-exercise-code"}
Sys.setlocale(category = 'LC_ALL','en_US.UTF-8')
library(readr)
library(dplyr)
dataSet <- read_csv("http://www.jeroenclaes.be/r_crashcourse_for_linguists/datasets/class3_claes_2017.csv")
```

```{r ex="proportionsTypical.Action.Chain.Pos", type="sample-code"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column Typical.Action.Chain.Pos to a factor

# Compute a table for the Typical.Action.Chain.Pos variable, store it in the variable 'Typical.Action.Chain.PosTab'

# Compute proportions for that table

# Print 'Typical.Action.Chain.PosTab'


```

```{r ex="proportionsTypical.Action.Chain.Pos", type="solution"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column Typical.Action.Chain.Pos to a factor
dataSet <- dataSet %>%
  mutate(Typical.Action.Chain.Pos=as.factor(Typical.Action.Chain.Pos))
# Compute a table for the Typical.Action.Chain.Pos variable, store it in the variable 'Typical.Action.Chain.PosTab'
Typical.Action.Chain.PosTab<-table(dataSet$Typical.Action.Chain.Pos)
# Compute proportions for that table
Typical.Action.Chain.PosTab<-prop.table(Typical.Action.Chain.PosTab)
# Print 'Typical.Action.Chain.PosTab'
Typical.Action.Chain.PosTab
```

```{r ex="proportionsTypical.Action.Chain.Pos", type="sct"}
test_object("Typical.Action.Chain.PosTab",incorrect_msg="Make sure to compute the table!")
test_output_contains('Typical.Action.Chain.PosTab',  incorrect_msg = "Make sure to compute proportions for the table!")
success_msg("Great!")
```

### 3.2 Proportions for the `corpus` variable 
```{r ex="proportionscorpus1", type="pre-exercise-code"}
Sys.setlocale(category = 'LC_ALL','en_US.UTF-8')
library(readr)
library(dplyr)
dataSet <- read_csv("http://www.jeroenclaes.be/r_crashcourse_for_linguists/datasets/class3_claes_2017.csv")
```

```{r ex="proportionscorpus1", type="sample-code"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column corpus to a factor

# Compute a table for the corpus variable, store it in the variable 'corpusTab'

# Convert the corpusTab object to a data.frame 

# Compute proportions based on the Freq column of that data.frame. Store the proportions in a new column called 'prop'

# Print corpusTab

```

```{r ex="proportionscorpus1", type="solution"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column corpus to a factor
dataSet <- dataSet %>%
  mutate(corpus=as.factor(corpus))
# Compute a table for the corpus variable, store it in the variable 'corpusTab'
corpusTab<-table(dataSet$corpus)
# Convert 'corpusTab' to a data.frame
corpusTab<-as.data.frame(corpusTab)
# Compute proportions based on the Freq column of that data.frame. Store the proportionsin a new column called 'prop'
corpusTab$prop<-corpusTab$Freq/sum(corpusTab$Freq, na.rm=TRUE)
# Print corpusTab
corpusTab
```

```{r ex="proportionscorpus1", type="sct"}
test_object("corpusTab",incorrect_msg="Make sure to compute the table!")
test_output_contains('corpusTab',  incorrect_msg="Make sure to compute proportions for the table!")
success_msg("Great!")
```

## 4. Bar plots 
### 4.1 Exploring the counts of the `corpus` variable (1/2)
- **ANALYZE**:
   - Which corpus provided the most examples?

```{r ex="freq1", type="pre-exercise-code"}
Sys.setlocale(category = 'LC_ALL','en_US.UTF-8')
library(readr)
library(dplyr)
dataSet <- read_csv("http://www.jeroenclaes.be/r_crashcourse_for_linguists/datasets/class3_claes_2017.csv")
```

```{r ex="freq1", type="sample-code"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column corpus to a factor

# Print a table of the corpus factor 

```

```{r ex="freq1", type="solution"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column corpus to a factor
dataSet <- dataSet %>%
  mutate(corpus=as.factor(corpus))

# Print a table of the corpus factor 
table(dataSet$corpus)
```

```{r ex="freq1", type="sct"}
test_output_contains("table(dataSet$corpus)", incorrect_msg = "Don't forget to print the table!")
success_msg("Great!")
```

### 4.2 Exploring the counts of the `corpus` variable (2/2)
```{r ex="freq2", type="pre-exercise-code"}
Sys.setlocale(category = 'LC_ALL','en_US.UTF-8')
library(readr)
library(dplyr)
dataSet <- read_csv("http://www.jeroenclaes.be/r_crashcourse_for_linguists/datasets/class3_claes_2017.csv")
```

```{r ex="freq2", type="sample-code"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column corpus to a factor

# Load ggplot2

# Draw a basic bar chart of the corpus factor

```

```{r ex="freq2", type="solution"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column corpus to a factor
dataSet <- dataSet %>%
  mutate(corpus=as.factor(corpus))
# Load ggplot2
library(ggplot2)
# Draw a basic bar chart of the corpus factor
ggplot(dataSet, aes(x=corpus)) + geom_bar()
```

```{r ex="freq2", type="sct"}
test_library_function("ggplot2", incorrect_msg = "Don't forget to load ggplot2")
test_ggplot(index = 1)
success_msg("Great!")
```

### 4.3 Exploring the counts of the `broad.regions` variable (1/2)
- **ANALYZE**:
  - Which broad.regions provided the most examples?

```{r ex="broad.regions_1", type="pre-exercise-code"}
Sys.setlocale(category = 'LC_ALL','en_US.UTF-8')
library(readr)
library(dplyr)
dataSet <- read_csv("http://www.jeroenclaes.be/r_crashcourse_for_linguists/datasets/class3_claes_2017.csv")
```

```{r ex="broad.regions_1", type="sample-code"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column broad.regions to a factor

# Print a table of the broad.regions factor

```

```{r ex="broad.regions_1", type="solution"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column broad.regions to a factor
dataSet <- dataSet %>% 
  mutate(broad.regions= as.factor(broad.regions))

# Print a table of the broad.regions factor
table(dataSet$broad.regions)
```

```{r ex="broad.regions_1", type="sct"}
 test_output_contains('table(dataSet$broad.regions)',  incorrect_msg = "Make sure to print the table!")

success_msg("Great!")
```

### 4.3 Exploring the counts of the `broad.regions` variable (2/2)
```{r ex="broad.regions", type="pre-exercise-code"}
Sys.setlocale(category = 'LC_ALL','en_US.UTF-8')
library(readr)
library(dplyr)
dataSet <- read_csv("http://www.jeroenclaes.be/r_crashcourse_for_linguists/datasets/class3_claes_2017.csv")
```

```{r ex="broad.regions", type="sample-code"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column broad.regions to a factor

# Load ggplot2

# Draw a pretty bar chart of the broad.regions factor: each region has its own fill and color, there is no legend title, the Y axis title is "Number of occurrences", the X axis title is "Large regions", the plot title is "Number of occurrences per region"

```

```{r ex="broad.regions", type="solution"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column broad.regions to a factor
dataSet <- dataSet %>% 
  mutate(broad.regions= as.factor(broad.regions))

# Load ggplot2
library(ggplot2)
# Draw a basic bar chart of the broad.regions factor
ggplot(dataSet, aes(x=broad.regions, fill=broad.regions, color=broad.regions)) + geom_bar() + theme(legend.title = element_blank()) + labs(x="Large regions", y="Number of occurrences", title="Number of occurrences per region")
```

```{r ex="broad.regions", type="sct"}
test_library_function("ggplot2")
test_ggplot(index=1, all_fail_msg = "Computer says no. This plot is completely wrong", data_fail_msg = "Make sure you include the data in the ggplot call", aes_fail_msg = "Make sure you define the correct aesthetics (aes)", geom_fail_msg = "Make sure you define the correct plotting function (geom_)")
success_msg("Great!")
```



### 4.4 Exploring the proportions of the `corpus` variable
```{r ex="prop2", type="pre-exercise-code"}
Sys.setlocale(category = 'LC_ALL','en_US.UTF-8')
library(readr)
library(dplyr)
dataSet <- read_csv("http://www.jeroenclaes.be/r_crashcourse_for_linguists/datasets/class3_claes_2017.csv")
```

```{r ex="prop2", type="sample-code"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column corpus to a factor

# Load ggplot2

# Draw a basic bar chart of the proportions of the corpus factor

```

```{r ex="prop2", type="solution"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column corpus to a factor
dataSet <- dataSet %>%
  mutate(corpus=as.factor(corpus))
# Load ggplot2
library(ggplot2)
# Draw a basic bar chart of the proportions of the corpus factor
ggplot(dataSet, aes(x=corpus)) + geom_bar(aes(y=..count../sum(..count..)))
```

```{r ex="prop2", type="sct"}
test_library_function("ggplot2")
test_ggplot(index=1, all_fail_msg = "Computer says no. This plot is completely wrong", data_fail_msg = "Make sure you include the data in the ggplot call", aes_fail_msg = "Make sure you define the correct aesthetics (aes)", geom_fail_msg = "Make sure you define the correct plotting function (geom_)")

success_msg("Great!")
```

### 4.5 Exploring the proportions of the `broad.regions` variable (1/2)
- **ANALYZE**:
      - Which broad.regions provided the most examples?

```{r ex="broad.regions_1", type="pre-exercise-code"}
Sys.setlocale(category = 'LC_ALL','en_US.UTF-8')
library(readr)
library(dplyr)
dataSet <- read_csv("http://www.jeroenclaes.be/r_crashcourse_for_linguists/datasets/class3_claes_2017.csv")
```

```{r ex="broad.regions_1", type="sample-code"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column broad.regions to a factor

# Print a table of the proportions of the broad.regions factor 

```

```{r ex="broad.regions_1", type="solution"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column broad.regions to a factor
dataSet <- dataSet %>% 
  mutate(broad.regions= as.factor(broad.regions))

# Print a table of the proportions of the broad.regions factor 
table(dataSet$broad.regions)/sum(table(dataSet$broad.regions))
```

```{r ex="broad.regions_1", type="sct"}
 test_output_contains('table(dataSet$broad.regions)',  incorrect_msg = "Make sure to print the table!")

success_msg("Great!")
```

### 4.5 Exploring the proportions of the `broad.regions` variable (2/2)
```{r ex="broad.regions_prop", type="pre-exercise-code"}
Sys.setlocale(category = 'LC_ALL','en_US.UTF-8')
library(readr)
library(dplyr)
dataSet <- read_csv("http://www.jeroenclaes.be/r_crashcourse_for_linguists/datasets/class3_claes_2017.csv")
```

```{r ex="broad.regions_prop", type="sample-code"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column broad.regions to a factor

# Load ggplot2

# Draw a pretty bar chart of the proportions of the broad.regions factor: each region has its own fill and color, there is no legend title, the Y axis title is "Number of occurrences", the X axis title is "Large regions", the plot title is "Number of occurrences per region"

```

```{r ex="broad.regions_prop", type="solution"}
# The data.frame dataSet is already in your workspace. The dplyr package is already loaded

# Convert the column broad.regions to a factor
dataSet <- dataSet %>% 
  mutate(broad.regions= as.factor(broad.regions))
# Load ggplot2
library(ggplot2)
# Draw a pretty bar chart of the proportions of the broad.regions factor: each region has its own fill and color, there is no legend title, the Y axis title is "Number of occurrences", the X axis title is "Large regions", the plot title is "Number of occurrences per region"
ggplot(dataSet, aes(x=broad.regions, fill=broad.regions, color=broad.regions)) + geom_bar(aes(y=..count../sum(..count..))) + theme(legend.title = element_blank()) + labs(x="Large regions", y="Number of occurrences", title="Number of occurrences per region")
```

```{r ex="broad.regions_prop", type="sct"}
test_library_function("ggplot2")
test_ggplot(index=1, all_fail_msg = "Computer says no. This plot is completely wrong", data_fail_msg = "Make sure you include the data in the ggplot call", aes_fail_msg = "Make sure you define the correct aesthetics (aes)", geom_fail_msg = "Make sure you define the correct plotting function (geom_)")
success_msg("Great!")
```

## 5. References
- Claes, J. (2017). Cognitive and geographic constraints on morphosyntactic variation: The variable agreement of presentational haber in Peninsular Spanish. *Belgian Journal of Linguistics* (31), 28-53.